---
title: "Exploring Bacteria"
author: "Pavlov Alexander"
date: '20 декабря 2018 г '
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting ready

This time we are going to conduct study concerning bacteria number of colonies forming units (cfu's) of airborne bacteria in the air in and around Graz. 

```{r, warning=FALSE}
library("readxl")
library("RColorBrewer")
library("MASS")
library("lme4")
pal<-brewer.pal(8,"Set3")

```

# Access Data

You can find the data either in http://www.stat.tugraz.
at/courses/files/BacteriaData.xlsx (sheet bacteria) or by clicking on Bacteria
Data at http://www.stat.tugraz.at/courses/glmLjubljana.html (only for those of
you experiencing troubles with the use of read.xls from within R).
The data resulted from a one year study in which bacteria/fungi (actually the number of
colonies forming units, cfu's) in the outdoor air were monitored at 7 dierent sites that
can be characterized as follows:

(a) village zone, near big farms with liquid manure pits and dung-hills;
(b) grassland and arable land, without buildings;
(c) suburban area with one-family houses and small farms;
(d) busy crossing, near a slaughter-house;
(e) public park on top of the Schloberg in the center of Graz;
(f) living area with apartment buildings and gardens;
(g) as for (f) but with compost arrangements.

Let us load the data into R:

```{r}
setwd("C:/Users/pavlov-aa/Desktop/GLM Multilevel")
ds<-read_xlsx("bacteriaData.xlsx")
ds<-data.frame(ds)
```


# Data Manipulation

Now let's prepare dataset for analysis:
1) Date and Site variables to factors
2) Create dependent variables of interest: "bac" and "fun" as sums of bi and fi
3) Check the structure

```{r}
ds[,1]<-as.factor(ds[,1])
ds[,2]<-as.factor(ds[,2])

ds[,"bac"]<-ds[,"b1"]+ds[,"b2"]+ds[,"b3"]+ds[,"b4"]+ds[,"b5"]+ds[,"b6"]
ds[,"fun"]<-ds[,"f1"]+ds[,"f2"]+ds[,"f3"]+ds[,"f4"]+ds[,"f5"]+ds[,"f6"]

str(ds)
```

Now, structure seems to be okay.

# Linear Regression

After validating models with different parameters we can see 2 best:

```{r ols}
lm6<-lm(data=ds, bac~temp+site)
lm7<-lm(data=ds, bac~humi+temp+site)
summary(lm6)
summary(lm7)
mylm<-lm6
```

lm6 model has a bit less R2, but all the variable are significant and higher R2 adjusted. I think its more safe to drop "humi" and count lm6 as the best linear model without transforming variables.

```{r plotting best linear model}
plot(lm6)
```

Normal Q-Q plot shows us that actually there are many observations with high error.
Taking into account, that R2 adjusted is about 30% we can say that model is weak.

Also lets look at interdependency between dependent and independent variables:

```{r echo=FALSE}
plot(ds$humi, ds$bac, xlab="Humidity", ylab="Bacteria colonies",main="Bacteria colines versus Humidity",col=pal); lines(lowess(ds$humi, ds$bac))
plot(ds$temp, ds$bac, xlab="Temperature", ylab="Bacteria colonies",main="Bacteria colines versus Temperature",col=pal); lines(lowess(ds$temp, ds$bac))
boxplot(ds$bac~ds$site, xlab="Site", ylab="Bacteria colonies",main="Bacteria colines versus Site",col=pal)
```

Humidity doesn't help with explaining Bacteria colonies. 
More temperature - more colonies and more variation. So, heterogenity exists.
In some places the ditribution may distinguish significantly.

And what about humi and temp of degree 2?

```{r quadro}
lm9<-lm(data=ds, bac~humi+temp+site+I(humi^2)+I(temp^2))
summary(lm9)
```

Better results but still R2 adjusted is not high. And suddenly humi became significant. But its not good. Plot showed that humi is uninformative.

What about assumptions?

``` {r assumptions, echo=FALSE}
hist(ds$bac,breaks=30,col=pal,xlab="Bacteria colonies",main="Histogram of colonies")
hist(ds$temp,breaks=30,col=pal,xlab="Temperature",main="Histogram of Temperature")
hist(ds$humi,breaks=30,col=pal,xlab="Humidity",main="Boxplot of Humidity")
```

Plot show us, that the distribution is not normal. So, assumptions are not met.
And what about residuals?

```{r errors, echo=FALSE}
plot(ds$bac, residuals(mylm),xlab="Bacteria colonies", ylab="residuals",main="Residuals of mylm"); abline(0, 0)
lines(lowess(ds$bac, residuals(mylm))) 
```

Bad situation. Maybe Box-Cox will help?

# Box-Cox Transformation

```{r bc}

bc<-boxcox(ds$bac~ds$humi+ds$temp+ds$site,lambda=seq(0.0,0.6,length=100),plotit=T)
ml.index <- which(bc$y == max(bc$y))
bc$x[ml.index]
lambda<-1/4

mylmBC<-lm(data=ds, ((bac^lambda)-1)/lambda~humi+temp+site)

plot(ds$bac, residuals(mylmBC), ylab="residuals"); abline(0, 0)

summary(mylmBC)
summary(mylm)

```

With Box-Cox transformation it became a bit better. With lambda =1 and =0 its worse.

```{r}
lambda<-1
mylmBC2<-lm(data=ds, ((bac^lambda)-1)/lambda~humi+temp+site)
mylmBC3<-lm(data=ds, log(bac)~humi+temp+site)
summary(mylmBC) # best
summary(mylmBC2) # poor
summary(mylmBC3) # poor
plot(mylmBC)
```

# GLM
```{r glm}
myglm<-glm(data=ds,bac~temp+site, family=gaussian(link="log"))
myglm2<-glm(data=ds,bac~temp+site,family=Gamma(link="log"))

summary(myglm)
plot(myglm)
plot(mylm)

plot(ds$bac, residuals(mylm), ylab="residuals"); abline(0, 0)
lines(lowess(ds$bac, residuals(mylm))) 
lines(lowess(ds$bac, residuals(myglm)),col="red") 
lines(lowess(ds$bac, residuals(myglm2)),col="blue") 

plot(myglm2)
```

Gamma is great. Residuals became smaller.

Prediction regions: site=6, humi=0.6
```{r pr}
m1<-min(ds$temp)
m2<-max(ds$temp)
temp<-seq(m1,m2,1)
site<-rep(6,38)
test<-data.frame(temp,"site"=as.factor(site))

p1<-predict.glm(myglm,test)
p2<-predict.glm(myglm2,test)

plot(temp,p1,main="Prediction region of colonies vs temp (points-glm1/red line-glm2)")
lines(temp,p2,col="red")
```

# Multilevel Models

```{r m}

lambda<-1/4
ds$bac_bc<-((ds$bac^lambda)-1)/lambda
hist(ds$bac_bc,breaks=30,col=pal,xlab="Bacteria colonies after Box-Cox",main="Histogram of colonies after Box-Cox") # okay seems to be normal


mlmod<- glmer(I(bac_bc+0.0001) ~ site+(temp+humi|date),data=ds,family=gaussian(link="log"))
mlmod2<- glmer(I(bac_bc+0.0001) ~ site+(temp+humi|date),data=ds, family=Gamma(link="log"))

summary(mlmod)
summary(mlmod2)

```

# AIC of all models:

```{r}

AIC(mylm)
AIC(mylmBC)
AIC(myglm)
AIC(myglm2)
AIC(mlmod) # best
AIC(mlmod2)

```

# Conclusion

In this case Box-Cox and Multilevel modelling gave us better results. Final diagnostics for mlmod model:

```{r dia}
plot(mlmod)
```

The results are similar to Box-Cox transformation.
